---
title: '"Chronic Kidney Disease"'
author: "Courtney D. Shelley, PhD"
date: "2026-07-14"
output: html_document
---

```{r sessionSetup, echo = FALSE, message = FALSE}
library(here)
```

### INTRODUCTION

This demo analysis will explore the Chronic Kidney Disease dataset [doi:
10.24432/C5G020](https://archive.ics.uci.edu/dataset/336/chronic+kidney+disease) available from the UC Irvine Machine Learning Repository. The dataset consists of 400 observations on 24 features, disease status, and an ID variable and can be used to predict chronic kidney disease in hospitalized patients. 

The 24 features include: 

* age		-	age	
* bp		-	blood pressure
* sg		-	specific gravity

### METHODOLOGY

#### Data Cleaning and Processing

Data was downloaded from the UCI Machine Learning Repository website as a .csv file. The dataset dimensions were verified. 

```{r loadData, include = FALSE}
## data read-in / verify
ckd = read.csv("kidney_disease.csv")
dim(ckd)
```

To prepare data for analysis, we first used pseudocode to explicitly state the data transformation pipeline with human-readable goals. 

```{r transform_1, include = FALSE}
# CLEAN CODE
# why:  Ensure informative variable names.
# what: Input a vector of column names. Output a vector of column names.
# how:  View names and replace if confusing

# why:  Ensure accurate variable values.
# what: Input a vector of column values. Output a vector of column values.
# how:  View and handle missingness
#       Ensure variable classes are accurate
#       Check category labels
```

We next transformed our pseudocode by identifying base R functions that can perform our necessary steps. 

```{r transform_2, include = FALSE}
# CLEAN CODE
# why:  Ensure informative variable names.
# what: Input a vector of column names. Output a vector of column names.
# how:  names()     # view names
#       colnames()  # replace names

# why:  Ensure accurate variable values.
# what: Input a vector of column values. Output a vector of column values.
# how:  apply() + perc_missingness # view rows/columns to handle missingness
#       class()                    # assess class accuracy
#       unique()                   # labels
```

We next performed the transformations and tested code at each step. In Step 1, we adjusted coded names into fully readable names. 

```{r transform_3, include = FALSE}
# CLEAN CODE
# why:  Ensure informative variable names.
# what: Input a vector of column names. Output a vector of column names.
# how:  names()     # view names
#       colnames()  # replace names

names(ckd)
colnames(ckd) <- c("id", "age", "blood_pressure", "specific_gravity", "albumin",
                   "sugar", "red_blood_cells", "pus_cell", "pus_cell_clumps",
                   "bacteria", "blood_glucose_random", "blood_urea", 
                   "serum_creatinine", "sodium", "potassium", "hemoglobin",
                   "packed_cell_volume", "white_blood_cell_count",
                   "red_blood_cell_count", "hypertension", "diabetes_mellitus",
                   "coronary_artery_disease", "appetite", "pedal_adema",
                   "anemia", "classification")
names(ckd) # test that names changed to long_form
```
In Step 2, we removed columns with a high level of missingness and verify that this was correctly performed. 

```{r transform_4, include = FALSE}
# CLEAN CODE
# why: ensure data accuracy
# what: clean_values()
# how: apply() + perc_missingness # view rows/columns to handle missingness

apply(ckd, 1, function(x) round(sum(is.na(x))/length(x)*100, 2))
apply(ckd, 2, function(x) round(sum(is.na(x))/length(x)*100, 2))

dim(ckd)
(which_sodium_potassium <- c(which(names(ckd) == "sodium"), which(names(ckd) == "potassium")))
ckd <- ckd[,-which_sodium_potassium]
ncol(ckd)   # test that we removed two columns

# handle missingness
sum(is.na(ckd))
ckd[ckd == ""] <- NA
sum(is.na(ckd))   # test that we increased number of NAs
```

In Step 3, we ensured data accuracy including recoding of some numeric variables to accurately reflect their variable type.

```{r transform_5, include = FALSE}
# CLEAN CODE
# why: ensure data accuracy
# what: clean_values()
# how: class()                    # assess class accuracy

str(ckd)

# $packed_cell_volume     : chr  "44" "38" "31" "32" ...
class(ckd$packed_cell_volume)
ckd$packed_cell_volume <- as.numeric(ckd$packed_cell_volume)
class(ckd$packed_cell_volume)     # test that class equals "numeric"

# $white_blood_cell_count : chr  "7800" "6000" "7500" "6700" ...
class(ckd$white_blood_cell_count)
ckd$white_blood_cell_count <- as.numeric(ckd$white_blood_cell_count)
class(ckd$white_blood_cell_count)  # test that class equals "numeric" 

# $$red_blood_cell_count   : chr  "5.2" "" "" "3.9" ...   
class(ckd$red_blood_cell_count)
ckd$red_blood_cell_count <- as.numeric(ckd$red_blood_cell_count)
class(ckd$red_blood_cell_count)    # test that class equals "numeric" 
```

In Step 4, we replaced variable value typos to ensure ensure accurate category labels. 

```{r transform_6, include = FALSE}
# CLEAN CODE
# why: ensure data accuracy
# what: clean_values()
# how: unique()                   # labels

apply(ckd, 2, unique)

# $diabetes_mellitus        [1] "yes"   "no"    " yes"  "\tno"  "\tyes" NA 
unique(ckd$diabetes_mellitus)
ckd$diabetes_mellitus[ckd$diabetes_mellitus == " yes"] <- "yes"
ckd$diabetes_mellitus[ckd$diabetes_mellitus == "\tno"] <- "no"
ckd$diabetes_mellitus[ckd$diabetes_mellitus == "\tyes"] <- "yes"
unique(ckd$diabetes_mellitus)   # test that all values are "yes", "no", or NA

# $coronary_artery_disease  [1] "no"   "yes"  "\tno" NA 
unique(ckd$coronary_artery_disease)
ckd$coronary_artery_disease[ckd$coronary_artery_disease == "\tno"] <- "no"
unique(ckd$coronary_artery_disease)  # test that all values are "yes", "no", or NA

# $classification           [1] "ckd"    "ckd\t"  "notckd"
unique(ckd$classification)
ckd$classification[ckd$classification == "ckd\t"] <- "ckd"
unique(ckd$classification)      # test that all values are "ckd" or "notckd"
```

## RESULTS

Our complete research workflow is presented below:
```{r mermaid, echo = FALSE}
library(DiagrammeR)

mermaid("
        graph LR
          A[Bronze] --> B[Silver]
          B --> C[Analysis]
          C --> E[Table]
          B --> D[Gold]
          C --> D
          D --> F[DAG]
          D --> G[Model]
          F --> H[Results]
        style A fill:#E68C47,stroke:#333,stroke-width:2px
        style B fill:#C9BFBD,stroke:#333,stroke-width:2px
        style D fill:#F2CB2E,stroke:#333,stroke-width:2px
        
        style E fill:#94F4D3,stroke:#333,stroke-width:2px
        style F fill:#94F4D3,stroke:#333,stroke-width:2px
        style G fill:#94F4D3,stroke:#333,stroke-width:2px
")
```

_Table 1: Comparability of Study Groups_

```{r gt_table, echo = FALSE, message = FALSE, warning = FALSE}
# CREATE TABLE 1
# why:  Summarize disease group comparability in a table 
# what: Input a prepared data.frame. Output a table.
# how:  


library(dplyr)
library(gt)
means <- ckd |>
            group_by(classification) |>
            summarize(across(where(is.numeric), list(mean = ~mean(.x, na.rm = TRUE))))

sds <- ckd |>
  group_by(classification) |>
  summarize(across(where(is.numeric), ~sd(.x, na.rm = TRUE))) 

ns <- ckd |>
          group_by(classification) |>
          summarize(across(where(is.numeric), list(n = ~sum(!is.na(.x)))))

names <- sub("_mean", "", names(means))
table1 = data.frame(Variable = names,
                    Mean_ckd = as.numeric(means[1,]),
                    sd_ckd = -as.numeric(sds[1,]),
                    n_ckd = as.numeric(ns[1,]),
                    perc_ckd = as.numeric(ns[1,])/sum(as.numeric(ns[1,]), na.rm = TRUE),
                    Mean_nockd = as.numeric(means[2,]),
                    sd_nockd = -as.numeric(sds[2,]),
                    n_nockd = as.numeric(ns[2,]), 
                    perc_nockd = as.numeric(ns[2,])/sum(as.numeric(ns[2,]), na.rm = TRUE))

table1 <- table1[3:14,]

table1 |> 
  gt() |>
    tab_header(
    title = md("**Table 1: Laboratory Value Comparisons**"),
    subtitle = "Chronic Kidney Disease (CKD) Patients and Non-CKD Patients") |>
    tab_stubhead(
      label = "Variable") |>
    tab_spanner(
      label = "Chronic Kidney Disease",
      columns = c(Mean_ckd, sd_ckd, n_ckd, perc_ckd)) |>
    tab_spanner(
      label = "No Disease",
      columns = c(Mean_nockd, sd_nockd, n_nockd, perc_nockd)) |>
    cols_label(
      Mean_ckd = html("Mean"),
      sd_ckd = html("sd"),
      n_ckd = html("n"),
      perc_ckd = html("(%)"),
      Mean_nockd = html("Mean"),
      sd_nockd = html("sd"),
      n_nockd = html("n"), 
      perc_nockd = html("(%)")) |>
  tab_style(
    style = cell_borders(
      sides = "left", 
      weight = px(1)), 
    locations = cells_body(
      columns = Mean_nockd)) |>
  tab_style(
    style = cell_borders(
      sides = "left",
      weight = px(2)), 
    locations = cells_body(
      columns = Mean_ckd)) |>
  tab_style(
    style = cell_text(style = "italic"),
    locations = cells_body(columns = c(sd_ckd, sd_nockd))) |>
   fmt_number(
    columns = c(Mean_ckd, sd_ckd, perc_ckd, Mean_nockd, sd_nockd, perc_nockd),
    decimals = 2) |>
  fmt_percent(
    columns = c(perc_ckd, perc_nockd)) |>
  fmt_number(
    columns = c(sd_ckd, sd_nockd),
    accounting = TRUE)
```


#### Analysis

We first examine a simple logistic regression model of the form: 

$$p_x = P(D|X = x) = \frac{1}{1+e^{-(a + bx)}}. $$
An alternative way of expressing this relationship is in terms of the log odds associated with $p_x$ as follows: 

$$log\left(\frac{p_x}{1-p_x} \right) = log(\text{odds for D|X = x}) = a + bx.$$

```{r load, echo = FALSE, warning = FALSE}

ckd$outcome = ckd$classification
ckd$outcome[ckd$outcome == "ckd"] <- 1
ckd$outcome[ckd$outcome == "notckd"] <- 0
ckd$outcome <- as.numeric(ckd$outcome)

# Fit the model
glm1 <- glm(outcome ~ serum_creatinine, data = ckd, 
            family = binomial)
summary(glm1)

creat_est = glm1$coef[[2]]
```

A model fitting ckd or no ckd as outcome with serum creatinine as predictor yields an estimate of `r creat_est` or an odds ratio of `r round(exp(creat_est), 3)`. 
