---
title: '"Chronic Kidney Disease"'
author: "Courtney D. Shelley, PhD"
date: "2026-07-14"
output: html_document
---

```{r sessionSetup}
library(here)
```

### INTRODUCTION

### METHODOLOGY

This demo analysis will explore the Chronic Kidney Disease dataset [doi:
10.24432/C5G020](https://archive.ics.uci.edu/dataset/336/chronic+kidney+disease) available from the UC Irvine Machine Learning Repository. The dataset consists of 400 observations on 24 features + chronic kidney disease diagnosis and can be used to predict chronic kidney disease in hospitalized patients. 

The 24 features include: 

* age		-	age	
* bp		-	blood pressure
* sg		-	specific gravity

#### Data Cleaning and Processing

We applied a medallion architecture to extract, transform, and load (ETL) source code. Medallion architecture performs ETL in three steps - a bronze layer archives raw data, a silver layer cleans data for general use, and a gold layer provides any model-specific transformation steps. 

```{r bronze}
## EXTRACT DATASET
# why:  Extract dataset from UCI Machine Learning Repository and archive
# what: Input dataset name. Output bronze dataset.
# how:  fetch_ucirepo()    # downloads dataset and metadata

#install.packages("ucimlrepo", repos = c('https://coatless-rpkg.r-universe.dev', 'https://cloud.r-project.org'))

library(ucimlrepo)

chronic_kidney_disease <- fetch_ucirepo(name = "Chronic Kidney Disease")

ckd_data = chronic_kidney_disease$data
X = ckd_data$features
Y = ckd_data$targets

ckd = cbind(X, Y)
class(ckd)  # test that class equals "data.frame"

dim(ckd)    # test that dimensions equals (400   25)
```

```{r silver_1}
# CLEAN CODE
# why:  Ensure informative variable names.
# what: Input a vector of column names. Output a vector of column names.
# how:  View names and replace if confusing

# why:  Ensure accurate variable values.
# what: Input a vector of column values. Output a vector of column values.
# how:  View and handle missingness
#       Ensure variable classes are accurate
#       Check category labels
```

For the silver layer, we first use pseudocode to explicitly state the data transformation pipeline with human-readable goals. 

```{r silver_2}
# CLEAN CODE
# why:  Ensure informative variable names.
# what: Input a vector of column names. Output a vector of column names.
# how:  names()     # view names
#       colnames()  # replace names

# why:  Ensure accurate variable values.
# what: Input a vector of column values. Output a vector of column values.
# how:  apply() + perc_missingness # view rows/columns to handle missingness
#       class()                    # assess class accuracy
#       unique()                   # labels
```

We next performed the transformations and tested code at each step. 

_(If we find we cannot follow our plan, we would return to step 2 and update the plan.)_

```{r silver3}
# CLEAN CODE
# why:  Ensure informative variable names.
# what: Input a vector of column names. Output a vector of column names.
# how:  names()     # view names
#       colnames()  # replace names

names(ckd)
colnames(ckd) <- c("id", "age", "blood_pressure", "specific_gravity", "albumin",
                   "sugar", "red_blood_cells", "pus_cell", "pus_cell_clumps",
                   "bacteria", "blood_glucose_random", "blood_urea", 
                   "serum_creatinine", "sodium", "potassium", "hemoglobin",
                   "packed_cell_volume", "white_blood_cell_count",
                   "red_blood_cell_count", "hypertension", "diabetes_mellitus",
                   "coronary_artery_disease", "appetite", "pedal_adema",
                   "anemia", "classification")
names(ckd) # test_that "names changed to long_form"
```

```{r silver_4}
# CLEAN CODE
# why: ensure data accuracy
# what: clean_values()
# how: apply() + perc_missingness # view rows/columns to handle missingness

apply(ckd, 1, function(x) round(sum(is.na(x))/length(x)*100, 2))
apply(ckd, 2, function(x) round(sum(is.na(x))/length(x)*100, 2))

dim(ckd)
(which_sodium_potassium <- c(which(names(ckd) == "sodium"), which(names(ckd) == "potassium")))
ckd <- ckd[,-which_sodium_potassium]
ncol(ckd)   # test_that "removed two columns"

# handle missingness
sum(is.na(ckd))
ckd[ckd == ""] <- NA
sum(is.na(ckd))   # test_that "increased number of NAs"
```

```{r silver_5}
# CLEAN CODE
# why: ensure data accuracy
# what: clean_values()
# how: class()                    # assess class accuracy

str(ckd)

# $packed_cell_volume     : chr  "44" "38" "31" "32" ...
class(ckd$packed_cell_volume)
ckd$packed_cell_volume <- as.numeric(ckd$packed_cell_volume)
class(ckd$packed_cell_volume)     # test_that "class equals "numeric" "

# $white_blood_cell_count : chr  "7800" "6000" "7500" "6700" ...
class(ckd$white_blood_cell_count)
ckd$white_blood_cell_count <- as.numeric(ckd$white_blood_cell_count)
class(ckd$white_blood_cell_count)  # test_that "class equals "numeric" "

# $$red_blood_cell_count   : chr  "5.2" "" "" "3.9" ...   
class(ckd$red_blood_cell_count)
ckd$red_blood_cell_count <- as.numeric(ckd$red_blood_cell_count)
class(ckd$red_blood_cell_count)    # test_that "class equals "numeric" "
```

```{r silver_6}
# CLEAN CODE
# why: ensure data accuracy
# what: clean_values()
# how: unique()                   # labels

apply(ckd, 2, unique)

# $diabetes_mellitus        [1] "yes"   "no"    " yes"  "\tno"  "\tyes" NA 
unique(ckd$diabetes_mellitus)
ckd$diabetes_mellitus[ckd$diabetes_mellitus == " yes"] <- "yes"
ckd$diabetes_mellitus[ckd$diabetes_mellitus == "\tno"] <- "no"
ckd$diabetes_mellitus[ckd$diabetes_mellitus == "\tyes"] <- "yes"
unique(ckd$diabetes_mellitus)   # test_that "all values are yes, no, or NA"

# $coronary_artery_disease  [1] "no"   "yes"  "\tno" NA 
unique(ckd$coronary_artery_disease)
ckd$coronary_artery_disease[ckd$coronary_artery_disease == "\tno"] <- "no"
unique(ckd$coronary_artery_disease)  # test_that "all values are yes, no, or NA"

# $classification           [1] "ckd"    "ckd\t"  "notckd"
unique(ckd$classification)
ckd$classification[ckd$classification == "ckd\t"] <- "ckd"
unique(ckd$classification)      # test_that "all values are ckd or notckd"
```

## RESULTS

Our complete research workflow is presented below:
```{r mermaid1}
library(DiagrammeR)

mermaid("
        graph LR
          A[Bronze] --> B[Silver]
          B --> C[Analysis]
          C --> E[Table]
          B --> D[Gold]
          C --> D
          D --> F[DAG]
          D --> G[Model]
          F --> H[Results]
        style A fill:#E68C47,stroke:#333,stroke-width:2px
        style B fill:#C9BFBD,stroke:#333,stroke-width:2px
        style D fill:#F2CB2E,stroke:#333,stroke-width:2px
        
        style E fill:#94F4D3,stroke:#333,stroke-width:2px
        style F fill:#94F4D3,stroke:#333,stroke-width:2px
        style G fill:#94F4D3,stroke:#333,stroke-width:2px
")
```

_Table 1: Comparability of Study Groups_

```{r gt_table}
# CREATE TABLE 1
# why:  Summarize disease group comparability in a table 
# what: Input a prepared data.frame. Output a table.
# how:  


library(dplyr)
library(gt)
means <- ckd |>
            group_by(classification) |>
            summarize(across(where(is.numeric), list(mean = ~mean(.x, na.rm = TRUE))))

sds <- ckd |>
  group_by(classification) |>
  summarize(across(where(is.numeric), ~sd(.x, na.rm = TRUE))) 

ns <- ckd |>
          group_by(classification) |>
          summarize(across(where(is.numeric), list(n = ~sum(!is.na(.x)))))

names <- sub("_mean", "", names(means))
table1 = data.frame(Variable = names,
                    Mean_ckd = as.numeric(means[1,]),
                    sd_ckd = -as.numeric(sds[1,]),
                    n_ckd = as.numeric(ns[1,]),
                    perc_ckd = as.numeric(ns[1,])/sum(as.numeric(ns[1,]), na.rm = TRUE),
                    Mean_nockd = as.numeric(means[2,]),
                    sd_nockd = -as.numeric(sds[2,]),
                    n_nockd = as.numeric(ns[2,]), 
                    perc_nockd = as.numeric(ns[2,])/sum(as.numeric(ns[2,]), na.rm = TRUE))

table1 <- table1[3:14,]

table1 |> 
  gt() |>
    tab_header(
    title = md("**Table 1: Laboratory Value Comparisons**"),
    subtitle = "Chronic Kidney Disease (CKD) Patients and Non-CKD Patients") |>
    tab_stubhead(
      label = "Variable") |>
    tab_spanner(
      label = "Chronic Kidney Disease",
      columns = c(Mean_ckd, sd_ckd, n_ckd, perc_ckd)) |>
    tab_spanner(
      label = "No Disease",
      columns = c(Mean_nockd, sd_nockd, n_nockd, perc_nockd)) |>
    cols_label(
      Mean_ckd = html("Mean"),
      sd_ckd = html("sd"),
      n_ckd = html("n"),
      perc_ckd = html("(%)"),
      Mean_nockd = html("Mean"),
      sd_nockd = html("sd"),
      n_nockd = html("n"), 
      perc_nockd = html("(%)")) |>
  tab_style(
    style = cell_borders(
      sides = "left", 
      weight = px(1)), 
    locations = cells_body(
      columns = Mean_nockd)) |>
  tab_style(
    style = cell_borders(
      sides = "left",
      weight = px(2)), 
    locations = cells_body(
      columns = Mean_ckd)) |>
  tab_style(
    style = cell_text(style = "italic"),
    locations = cells_body(columns = c(sd_ckd, sd_nockd))) |>
   fmt_number(
    columns = c(Mean_ckd, sd_ckd, perc_ckd, Mean_nockd, sd_nockd, perc_nockd),
    decimals = 2) |>
  fmt_percent(
    columns = c(perc_ckd, perc_nockd)) |>
  fmt_number(
    columns = c(sd_ckd, sd_nockd),
    accounting = TRUE)
```


#### Analysis

We first examine a simple logistic regression model of the form: 

$$p_x = P(D|X = x) = \frac{1}{1+e^{-(a + bx)}}. $$
An alternative way of expressing this relationship is in terms of the log odds associated with $p_x$ as follows: 

$$log\left(\frac{p_x}{1-p_x} \right) = log(\text{odds for D|X = x}) = a + bx.$$

```{r silver_to_gold}

ckd$outcome = ckd$classification
ckd$outcome[ckd$outcome == "ckd"] <- 1
ckd$outcome[ckd$outcome == "notckd"] <- 0
ckd$outcome <- as.numeric(ckd$outcome)

# Fit the model
glm1 <- glm(outcome ~ serum_creatinine, data = ckd, 
            family = binomial)
summary(glm1)

creat_est = glm1$coef[[2]]
```

A model fitting ckd or no ckd as outcome with serum creatinine as predictor yields an estimate of `r creat_est`. 
